Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-19 trunk/papers/ArticleZ80/source/paper.tex
Corrigido o formato. Estou resumindo o texto para as 16 páginas.

The new model supports animation in ProB~\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of the microcontroller model. Such
a library has many definitions of concepts common to microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using the Z80 assembly language to analyse the code simulation
and the verification process.

@@

\hspace*{0.0in}\it PHYS\_BYTE\_INDEX \rm = \rm 0 $\upto$ \rm (\it BYTE\_WIDTH\rm -\rm 1\rm )\hspace*{0.03in} $\land$ \hspace*{0.0in}\it BYTE \rm =\rm (\it BYTE\_INDEX  $\fun$  \it BIT\rm )\hspace*{0.0in} $\land$

@@

\item For the proof obligation (\it byte\_uchar\rm \ $\in$ \textit{BYTE} $\bij$
\textit{UCHAR}), the proof rule is:\\
$\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) \rm $\in$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )  $\bij$  \rm 0 $\upto$ \rm 2\rm 5\rm 5 \\

\item For the proof obligation (\it uchar\_byte\rm \ $\in$ \textit{UCHAR} $\bij$
\textit{BYTE}), the proof rule is:\\ $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 0$\upto$255  $\mid$  \rm \rm  [(\it v0  $\mod$  \rm 2)$\div$1,(\it v0  $\mod$  \rm 4)$\div$\rm 2\rm,(\it v0  $\mod$  \rm 8)$\div$\rm 4\rm,(\it v0  $\mod$  \rm 16)$\div$\rm 8\rm,(\it v0  $\mod$  \rm 32)$\div$\rm 1\rm 6\rm,(\it v0  $\mod$  \rm 6\rm 4)$\div$\rm 3\rm 2\rm,(\it v0  $\mod$  \rm 1\rm 2\rm 8)$\div$\rm 6\rm 4\rm,(\it v0  $\mod$  \rm 2\rm 5\rm 6)$\div$\rm 1\rm 2\rm 8\rm \rm ]\rm ) \rm   $\in$  \rm 0 $\upto$ \rm 2\rm 5\rm 5   $\bij$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )\\

@@

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller module Z80
and parts of it are presented below.  Basically, the state of microcontroller is
formed by data from program counter, ports, registers, etc. The transitions
between the states are actioned by execution of instruction or an external
action. Groups of registers are represented by the variables of the
specification states that appear in clause \textit{VARIABLES}. The declaration
of valid states of variables is in the \textit{INVARIANT} clause and the initial
state is defined in the \textit{INITIALISATION} clause. The assembly
instructions are defined by the clause \textit{OPERATIONS} and, in general, each
instruction has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
General functions that can be used in other microcontroller models are defined
in the modules of data types. Specific functions of the microcontroller are
defined in the ALU (arithmetic logic unit) module.

The main module includes an instance of the memory module and accesses the
definitions from basic data types modules and the \textit{ALU} module.

@@

\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ \it pc  $\in$  \it INSTRUCTION  $\land$  \it sp  $\in$  \it BV16 $\land$  \it ix  $\in$  \it BV16  $\land$

\hspace*{0.10in}\it iy  $\in$  \it BV16  $\land$ \it i\_  $\in$  \it BYTE  $\land$  \it r\_ $\in$  \it BYTE  $\land$ \it iff1  $\in$  \it BIT  $\land$ \it iff2  $\in$  \it BIT  $\land$

\hspace*{0.10in}\it im $\in$ (\it BIT $\times$ \it BIT\rm )  $\land$  \it i\_o\_ports  $\in$  \it BYTE  $\fun$  \it BYTE
\end{sloppypar}

@@

	\bf SETS\\
\hspace*{0.10in}\it id\_reg\_8 \rm = \rm \{ \it a0\rm, \it f0\rm, \it b0\rm, \it
c0\rm, \it d0\rm, \it e0\rm, \it h0\rm, \it l0\rm, \it a\_0\rm,  \it f\_0\rm,
\it b\_0\rm, \it c\_0\rm, \it d\_0\rm, \it e\_0\rm, \it h\_0\rm, \it l\_0 \rm
\};\\ \hspace*{0.10in}\it id\_reg\_16 \rm = \rm \{  \it AF \rm, \it BC\rm, \it
DE\rm, \it HL\rm, \it SP\rm\}

@@

The main working register of Z80 is the accumulator ($\textit{rgs8(a0)}$) used for arithmetic/logic, input/output and loading/storing operations.
Another important element is the ``f'' register ($\textit{rgs8(f0)}$),  that is
used as a \textbf{flag register}. This register uses only six bits to represent the execution result status of each instruction.
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning: %$\textit{bv\_get(rgs8(f0),0)}$ - the carry bit; $\textit{bv\_get(rgs8(f0),1)}$ - the add/subtract bit; $\textit{bv\_get(rgs8(f0),2)}$ - the parity or overflow bit; $\textit{bv\_get(rgs8(f0),4)}$ - the half carry bit; $\textit{bv\_get(rgs8(f0),6)}$ - the zero bit; and $\textit{bv\_get(rgs8(f0),7)}$ - the sign bit.
\begin{description}
  \item[$\textit{bv\_get(rgs8(f0),0)}$] - The Carry bit.
  \item[$\textit{bv\_get(rgs8(f0),1)}$] - The Add/Subtract bit.
  \item[$\textit{bv\_get(rgs8(f0),2)}$] - The Parity or Overflow bit.
  \item[$\textit{bv\_get(rgs8(f0),4)}$] - The Half Carry bit.
  \item[$\textit{bv\_get(rgs8(f0),6)}$] - The Zero bit.
  \item[$\textit{bv\_get(rgs8(f0),7)}$] - The Sign bit.
\end{description}
These bits can also be used to specify security properties for microcontroller programs.

\textbf{Assuring the absence of overflow:}
 \emph{To assure that an overflow does not happen, the developer can add this
 expression (\textit{bv\_get(rgs8(f0),0)} $\neq$ 1 $\land$
 \textit{bv\_get(rgs8(f0),2) $\neq$ 1}) in the invariant.
% When a overflow happen,it can be dangerous. So the developer may restrict its use. However, this restriction can also
% become more difficult to verify the model.
}

@@

\subsection{Modeling the actions and instructions}

Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has three categories of operations: a category represents the 
instructions of microcontrollers, a second category represents the input and output of data 
(shown in ~\ref{sec:modelingIO}) and the last category represents the external
actions (shown in~\ref{sec:externalactions}). A simple example of instruction is
$\textit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The B tools does not allow
to use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown below. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module and it receives
a address memory and its new memory value. It also increments the program
counter ($\textit{pc}$) and update the refresh register ($\textit{r\_}$).
The other instructions have a similar structure.

@@

\subsection{Modeling the input/output instructions}
\label{sec:modelingIO}
The Z80 has an extensive set of input and output (I/O) instructions and 256 ports for
devices. This model can transfer data blocks between the I/O devices and any
of the internal registers or memory address.

The $\textit{IN\_r(C)}$\footnoterecall{myfootnote} instruction is represented
by the following B operation. It receives a register identifier ``$\textit{rr}$'' and, it stores the value from ``rr'' in the $\textit{C}$ port address. Besides, it increments the program counter and updates the flag registers.

\hspace*{0.0in}\bf IN\_r\_9C0 \rm ( \it rr \rm ) \rm =

\hspace*{0.0in}\bf PRE \it rr  $\in$  \it id\_reg\_8  $\land$  \it rr  $\not =$  \it f0\hspace*{0.15in}\bf THEN

\hspace*{0.20in}\bf ANY \hspace*{0.10in}\it negative \rm , \it zero \rm , \it half\_carry \rm , \it pv \rm , \it add\_sub \rm , \it carry

\hspace*{0.20in}\bf WHERE

\hspace*{0.40in}\it negative $\in$ \it BIT $\land$ \it zero $\in$ \it BIT $\land$ \it half\_carry $\in$ \it BIT 
$\land$ \it pv $\in$ \it BIT $\land$

 \hspace*{0.40in}\it add\_sub $\in$ \it BIT $\land$ \it carry $\in$ \it BIT  $\land$

\hspace*{0.40in}\it negative \rm = \it is\_negative \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$

\hspace*{0.40in}\it zero \rm = \it is\_zero \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$ \it half\_carry \rm = \rm 0  $\land$

\hspace*{0.40in}\it pv \rm = \it parity\_even \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) $\land$
\it add\_sub \rm =\hspace*{0.10in}\rm 0  $\land$ 
\it carry \rm = \it z\_c

\hspace*{0.20in}\bf THEN

\hspace*{0.40in}\it rgs8 \rm := \it rgs8  $\lover$  \rm \{ \rm ( \it rr  $\mapsto$  \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) \rm ,

\hspace*{0.40in}\it update\_flag\_reg\rm (\it negative\rm,\it zero\rm,\it half\_carry\rm,\it pv\rm,\it add\_sub\rm,\it carry)\rm\}$\para$

\hspace*{0.40in}\it pc \rm := \it instruction\_next \rm ( \it pc \rm )  $\para$  \it r\_ \rm := \it update\_refresh\_reg\rm (\it r\_\rm )

\hspace*{0.20in}\bf END

\hspace*{0.0in}\bf END \rm

%The main I/O instructions work similarly, for example: $\textit{OUT(n),A}$  or $\textit{OUT (C ), r}$.

% $
% \begin{array}{l} 
% \textit{IN\_r\_9C0} ( rr ) = \\
% \quad   \PRE rr \in id\_reg\_8    \THEN\\
% \quad\quad \ANY data\_in, negative , zero , half\_carry , pv , add\_sub , carry\\ 
% \quad\quad \WHERE data\_in \in \textit{BYTE} \land negative \in \textit{BIT}\land\\
% \quad\quad\quad carry \in \textit{BIT} \land half\_carry \in \textit{BIT} \land zero \in \textit{BIT} \land \\
% \quad\quad\quad negative = is\_negative (data\_in) \land zero = is\_zero(data\_in ) \land\\
% \quad\quad\quad half\_carry = 0 \land pv =parity\_even\_BYTE ( data\_in )    \land\\
% \quad\quad\quad add\_sub =  0 \land carry = z\_c \\
% \quad\quad  \THEN \\
% \quad\quad\quad i\_o\_ports ( rgs8 ( c0 ) ) := data\_in ||\\
% \quad\quad\quad rgs8 := rgs8 <+ \{ ( rr |-> data\_in ) ,\\
% \quad\quad\quad get\_new\_flag\_register\_SZ\_H\_PvNC ( rgs8 , negative , zero, half\_carry , pv , add\_sub , carry ) \} ||\\
% \quad\quad\quad pc := instruction\_next( pc )\\
% \quad\quad \END\\
% \quad \END\\
% \end{array}
% $

\subsection{Modeling the external actions}
\label{sec:externalactions}

The external actions change the state of the microcontroller, for example,
refreshing the I/O ports and interruption request. The external actions are also
modeled by operations and they are named with the prefix ``$ext\_$'' and
followed by the name of action. There are four external actions:
$ext\_update\_io\_ports$, $ext\_NMI$ and $ext\_INT$, $ext\_Reset$. The
instruction $ext\_update\_io\_ports$ just changes the state of I/O port, see.

@@

The other external actions are related to interruptions. Interruptions allow
devices to suspend a routine from CPU and start another service routine.
This service routine can exchange data or signals between CPU and external
devices. When a routine is finished, then the CPU comes back to the last routine
that was interrupted.

For the interrupts, the following elements are important:  the interrupt flip-flops
($\textit{iff1}$ and $\textit{iff2}$), the types of interrupts (maskable and
non-maskable), the interrupt mode (set with the $\textit{IM 0}$, $\textit{IM 1}$,
$\textit{IM 2}$ instructions) and the $\textit{i\_}$ register.

Flip-flops $\textit{iff1}$ and $\textit{iff2}$ control the maskable interrupts
($\textit{INT}$). When $\textit{iff1}$ is set, the interrupt is enabled,
otherwise it is disabled; $\textit{iff2}$ is used only as backup for $\textit{iff1}$. The
instructions $\textit{EI}$ and $\textit{DI}$ respectively enable and disable the
maskable interruptions, setting  $\textit{iff1}$ to 1 and 0.

The interruptions and the \textit{reset} action can change the state of program
counter. Theses actions are modeled by B operations and its main effects are
presented here\footnote{Some definitions of constants: $\textit{sp\_minus\_two}$
holds the value of stack pointer minus 2,
% =$\textit{dec\_BV16(dec\_BV16(sp))}$
 $\textit{sp\_minus\_one}$ is the value of stack pointer minus 1,
% =$\textit{dec\_BV16(sp)}$
$\textit{pc\_high}$ holds the most significant 8 bits and $\textit{pc\_low}$
holds the least significant 8 bits.}.

 \textbf{NMI} - Non-maskable interrupts cannot be disabled
 by the programmer. Then, when a device makes a request, $sp$ is pushed,
 $pc$ receives $66H$ (102 in decimal), $\textit{iff1}$ is reset, $\textit{iff2}$ stores
 $\textit{iff1}$ and the refresh register is updated.

\begin{sloppypar}
\bf updateStack\rm (\rm \{ \rm (\it sp\_minus\_two  $\mapsto$  \it pc\_low\rm )\rm ,\rm (\it sp\_minus\_one  $\mapsto$ \it pc\_high \rm ) \rm \}\rm )$\para$

\it sp \rm := \it sp\_minus\_two  $\para$ \it pc \rm := \rm 1\rm 0\rm 2 $\para$ \it iff1\rm :=\rm 0  $\para$  \it iff2\rm := \it iff1 $\para$

\it r\_ \rm := \it update\_refresh\_reg\rm (\it r\_\rm )\\
\end{sloppypar}

  \textbf{INT} - Maskable Interrupt is usually reserved for important functions
  that can be enabled and disabled by the programmer. When a maskable interrupt
  action happens, both $\textit{iff1}$ and $\textit{iff2}$ are cleared,
  disabling the interrupts, $sp$ is pushed, the refresh register is updated and
  the other effects depend on the interrupt mode register (\it im\rm).

 \begin{itemize}

  \item The mode 0 is compatible with 8080 and this mode is selected when \it im
  \rm = \rm ( \rm 0 $\mapsto$  \rm 0 \rm ). When a  non-maskable interrupt
  happens, the CPU fetches an instruction of one byte from an external device,
  usually an RST instruction, and the CPU executes it.
  The instruction code is received from an external device by data bus and it is
  represented by integer parameter called $\textit{byte\_bus}$.

  \item The mode 1 is the simplest and this mode is selected when \it im \rm =
  \rm ( \rm 0  $\mapsto$  \rm 1 \rm ). Simply, when a non-maskable interruption
  happens, the program counter receives $38H$ (56 in decimal).

  \item The mode 2 is the most flexible and this mode is selected when \it im
  \rm = \rm ( \rm 1  $\mapsto$ \rm 1 \rm ). When a non-maskable interruption
  happens, an indirect call can be made to any address memory. The program
  counter receives a bit vector of size 16 composed with two parts: the most
  significant part of the $\textit{i\_}$ register and the least significant part
  of the $\textit{byte\_bus}$ with the last bit cleared.

 \end{itemize}

The essential part of maskable interrupt is shown below, where
$\textit{byte\_bus}$ is a parameter of the $\textit{INT}$ operation:

\hspace*{0.00in}\bf IF \it im \rm = \rm ( \rm 0  $\mapsto$  \rm 0 \rm ) \bf THEN

>>>>

The new model supports animation in ProB~\cite{proB}.  Essentially, the
new model \footnote{The interested reader in more details is invited to visit our repository at: http://code.google.com/p/b2asm/.}
no longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of the microcontroller model. Such
a library has many definitions of concepts common to microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using the Z80 assembly language to analyse the code simulation
and the verification process.

@@

\hspace*{0.0in}\it PHYS\_BYTE\_INDEX \rm = \rm 0 $\upto$ \rm (\it BYTE\_WIDTH\rm -\rm 1\rm )\hspace*{0.03in} $\land$

\hspace*{0.0in}\it BYTE \rm =\rm (\it BYTE\_INDEX  $\fun$  \it BIT\rm )\hspace*{0.0in} $\land$

@@

\item For the proof obligation (\it byte\_uchar\rm \ $\in$ \textit{BYTE} $\bij$
\textit{UCHAR}), the proof rule is: $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) \rm $\in$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )  $\bij$  \rm 0 $\upto$ \rm 2\rm 5\rm 5 \\

\item For the proof obligation (\it uchar\_byte\rm \ $\in$ \textit{UCHAR} $\bij$
\textit{BYTE}), the proof rule is: $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 0$\upto$255  $\mid$  \rm \rm  [(\it v0  $\mod$  \rm 2)$\div$1,(\it v0  $\mod$  \rm 4)$\div$\rm 2\rm,(\it v0  $\mod$  \rm 8)$\div$\rm 4\rm,(\it v0  $\mod$  \rm 16)$\div$\rm 8\rm,(\it v0  $\mod$  \rm 32)$\div$\rm 1\rm 6\rm,(\it v0  $\mod$  \rm 6\rm 4)$\div$\rm 3\rm 2\rm,(\it v0  $\mod$  \rm 1\rm 2\rm 8)$\div$\rm 6\rm 4\rm,(\it v0  $\mod$  \rm 2\rm 5\rm 6)$\div$\rm 1\rm 2\rm 8\rm \rm ]\rm ) \rm   $\in$  \rm 0 $\upto$ \rm 2\rm 5\rm 5   $\bij$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )\\

@@

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller module Z80
and parts of it are presented below.  Basically, the state of microcontroller is
formed by data from program counter, ports, registers, etc. The transitions
between the states are actioned by execution of instruction or an external
action. Groups of registers are represented by the variables of the
specification states that appear in clause \textit{VARIABLES}. The declaration
of valid states of variables is in the \textit{INVARIANT} clause and the initial
state is defined in the \textit{INITIALISATION} clause. The assembly
instructions are defined by the clause \textit{OPERATIONS} and, in general, each
instruction has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
General functions that can be used in other microcontroller models are defined
in the modules of data types. Specific functions of the microcontroller are
defined in the ALU (arithmetic logic unit) module.The main module includes an instance of the memory module and accesses the
definitions from basic data types modules and the \textit{ALU} module.

@@

\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ \it pc  $\in$  \it INSTRUCTION  $\land$  \it sp  $\in$  \it BV16 $\land$

\hspace*{0.10in}\it ix  $\in$  \it BV16  $\land$ \it iy  $\in$  \it BV16  $\land$ \it i\_  $\in$  \it BYTE  $\land$  \it r\_ $\in$  \it BYTE  $\land$ \it iff1  $\in$  \it BIT  $\land$

\hspace*{0.10in}\it iff2  $\in$  \it BIT  $\land$ \it im $\in$ (\it BIT $\times$ \it BIT\rm )  $\land$  \it i\_o\_ports  $\in$  \it BYTE  $\fun$  \it BYTE
\end{sloppypar}

@@

\bf SETS\\
\hspace*{0.10in}\it id\_reg\_8\rm =\rm \{\it a0\rm, \it f0\rm, \it b0\rm, \it
c0\rm, \it d0\rm, \it e0\rm, \it h0\rm, \it l0\rm, \it a\_0\rm,  \it f\_0\rm,
\it b\_0\rm, \it c\_0\rm, \it d\_0\rm, \it e\_0\rm, \it h\_0\rm, \it l\_0 \rm
\};\\ \hspace*{0.10in}\it id\_reg\_16\rm =\rm \{\it AF \rm, \it BC\rm, \it
DE\rm, \it HL\rm, \it SP\rm\}

@@

The main working register of Z80 is the accumulator ($\textit{rgs8(a0)}$) used for arithmetic/logic, input/output and loading/storing operations.
Another important element is the ``f'' register ($\textit{rgs8(f0)}$),  that is
used as a \textbf{flag register}. This register uses only six bits to represent the execution result status of each instruction.
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning: $\textit{bv\_get(rgs8(f0),0)}$ - the carry bit; $\textit{bv\_get(rgs8(f0),1)}$ - the add/subtract bit; $\textit{bv\_get(rgs8(f0),2)}$ - the parity or overflow bit; $\textit{bv\_get(rgs8(f0),4)}$ - the half carry bit; $\textit{bv\_get(rgs8(f0),6)}$ - the zero bit; and $\textit{bv\_get(rgs8(f0),7)}$ - the sign bit.
%\begin{description}
%  \item[$\textit{bv\_get(rgs8(f0),0)}$] - The Carry bit.
%  \item[$\textit{bv\_get(rgs8(f0),1)}$] - The Add/Subtract bit.
%  \item[$\textit{bv\_get(rgs8(f0),2)}$] - The Parity or Overflow bit.
%  \item[$\textit{bv\_get(rgs8(f0),4)}$] - The Half Carry bit.
%  \item[$\textit{bv\_get(rgs8(f0),6)}$] - The Zero bit.
%  \item[$\textit{bv\_get(rgs8(f0),7)}$] - The Sign bit.
%\end{description}
These bits can also be used to specify security properties for microcontroller programs.

\textbf{Assuring the absence of overflow:}
 \emph{To assure that an overflow does not happen, the developer can add in the invariant this 
 expression:\\(\textit{bv\_get(rgs8(f0),0)} $\neq$ 1 $\land$
 \textit{bv\_get(rgs8(f0),2) $\neq$ 1}).
% When a overflow happen,it can be dangerous. So the developer may restrict its use. However, this restriction can also
% become more difficult to verify the model.
}

@@

\subsection{Modeling the actions and instructions}
%TODO: Fiz uma breve alteração
Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has two categories of operations: a category represents the 
instructions of microcontrollers and a category represents the external actions, e.g.
interruptions actions. All 158 instructions and interruptions were modeled, but only
two important examples are presented below.

A simple example of instruction is $\textit{LD\_(nn)\_A}$\footnote{The B tools does not allow o use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown below. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module, it receives
an address memory and its new memory value. $\textit{LD\_(nn)\_A}$ also increments the program
counter ($\textit{pc}$) and updates the refresh register ($\textit{r\_}$). The other instructions have a similar structure.

@@

%\subsection{Modeling the input/output instructions}
%\label{sec:modelingIO}
%%The Z80 has an extensive set of input and output (I/O) instructions and 256 ports for
%%devices. This model can transfer data blocks between the I/O devices and any
%%of the internal registers or memory address. The $\textit{IN\_r(C)}$ instruction is
%%represented by the following B operation. It receives a register identifier ``$\textit{rr}$''
%%and, it stores the value from ``rr'' in the $\textit{C}$ port address. Besides,
%%it increments the program counter and updates the flag registers.
%%
%%\hspace*{0.0in}\bf IN\_r\_9C0 \rm ( \it rr \rm ) \rm =
%%
%%\hspace*{0.0in}\bf PRE \it rr  $\in$  \it id\_reg\_8  $\land$  \it rr  $\not =$  \it f0\hspace*{0.15in}\bf THEN
%%
%%\hspace*{0.20in}\bf ANY \hspace*{0.10in}\it negative \rm , \it zero \rm , \it half\_carry \rm , \it pv \rm , \it add\_sub \rm , \it carry
%%
%%\hspace*{0.20in}\bf WHERE 
%%
%%\hspace*{0.40in}\it negative $\in$ \it BIT $\land$ \it zero $\in$ \it BIT $\land$ \it half\_carry $\in$ \it BIT 
%%$\land$ \it pv $\in$ \it BIT $\land$
%%
%% \hspace*{0.40in}\it add\_sub $\in$ \it BIT $\land$ \it carry $\in$ \it BIT  $\land$
%%
%%\hspace*{0.40in}\it negative \rm = \it is\_negative \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$ 
%%
%%\hspace*{0.40in}\it zero \rm = \it is\_zero \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$ \it half\_carry \rm = \rm 0  $\land$ 
%%
%%\hspace*{0.40in}\it pv \rm = \it parity\_even \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) $\land$
%%\it add\_sub \rm =\hspace*{0.10in}\rm 0 % $\land$ \it carry \rm = \it z\_c
%%
%%\hspace*{0.20in}\bf THEN
%%
%%\hspace*{0.40in}\it rgs8 \rm := \it rgs8  $\lover$  \rm \{ \rm ( \it rr  $\mapsto$  \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) \rm ,
%%
%%\hspace*{0.40in}\it update\_flag\_reg\rm (\it negative\rm,\it zero\rm,\it half\_carry\rm,\it pv\rm,\it add\_sub\rm,\it carry)\rm\}$\para$
%%
%%\hspace*{0.40in}\it pc \rm := \it instruction\_next \rm ( \it pc \rm )  $\para$  \it r\_ \rm := \it update\_refresh\_reg\rm (\it r\_\rm )
%%
%%\hspace*{0.20in}\bf END
%%
%%\hspace*{0.0in}\bf END \rm
%%
%%%The main input and output instructions work similarly, for example: $\textit{OUT(n),A}$  or $\textit{OUT (C ), r}$. 
%%
%% $
%% \begin{array}{l} 
%% \textit{IN\_r\_9C0} ( rr ) = \\
%% \quad   \PRE rr \in id\_reg\_8    \THEN\\
%% \quad\quad \ANY data\_in, negative , zero , half\_carry , pv , add\_sub , carry\\ 
%% \quad\quad \WHERE data\_in \in \textit{BYTE} \land negative \in \textit{BIT}\land\\
%% \quad\quad\quad carry \in \textit{BIT} \land half\_carry \in \textit{BIT} \land zero \in \textit{BIT} \land \\
%% \quad\quad\quad negative = is\_negative (data\_in) \land zero = is\_zero(data\_in ) \land\\
%% \quad\quad\quad half\_carry = 0 \land pv =parity\_even\_BYTE ( data\_in )    \land\\
%% \quad\quad\quad add\_sub =  0 \land carry = z\_c \\
%% \quad\quad  \THEN \\
%% \quad\quad\quad i\_o\_ports ( rgs8 ( c0 ) ) := data\_in ||\\
%% \quad\quad\quad rgs8 := rgs8 <+ \{ ( rr |-> data\_in ) ,\\
%% \quad\quad\quad get\_new\_flag\_register\_SZ\_H\_PvNC ( rgs8 , negative , zero, half\_carry , pv , add\_sub , carry ) \} ||\\
%% \quad\quad\quad pc := instruction\_next( pc )\\
%% \quad\quad \END\\
%% \quad \END\\
%% \end{array}
%% $
%
%
%
%
%%\subsection{Modeling the external actions}
%%\label{sec:externalactions}
%
The external actions change the state of the microcontroller, for example,
refreshing the I/O ports and requesting interruptions. The external actions are also
modeled by operations and they are named with the prefix ``$ext\_$'' and
followed by the name of action. There are four external actions:
$ext\_update\_io\_ports$, $ext\_NMI$ and $ext\_INT$, $ext\_Reset$. The
instruction $ext\_update\_io\_ports$ just changes the state of I/O port, see.

@@

The other external actions are related to interruptions. Interruptions allow
devices to suspend a routine from CPU and start another service routine.
This service routine can exchange data or signals between CPU and external
devices. When a routine is finished, then the CPU comes back to the last routine
that was interrupted. Modeling details about interrupts can be consulted in
public project's repository.

%
%For the interrupts, the following elements are important:  the interrupt flip-flops
%($\textit{iff1}$ and $\textit{iff2}$), the types of interrupts (maskable and
%non-maskable), the interrupt mode (set with the $\textit{IM 0}$, $\textit{IM 1}$,
%$\textit{IM 2}$ instructions) and the $\textit{i\_}$ register.
%
%Flip-flops $\textit{iff1}$ and $\textit{iff2}$ control the maskable interrupts
%($\textit{INT}$). When $\textit{iff1}$ is set, the interrupt is enabled,
%otherwise it is disabled; $\textit{iff2}$ is used only as backup for $\textit{iff1}$. The
%instructions $\textit{EI}$ and $\textit{DI}$ respectively enable and disable the
%maskable interruptions, setting  $\textit{iff1}$ to 1 and 0.
%
%
%The interruptions and the \textit{reset} action can change the state of program
%counter. Theses actions are modeled by B operations and its main effects are
%presented here\footnote{Some definitions of constants: $\textit{sp\_minus\_two}$
%holds the value of stack pointer minus 2,
%% =$\textit{dec\_BV16(dec\_BV16(sp))}$
% $\textit{sp\_minus\_one}$ is the value of stack pointer minus 1,
%% =$\textit{dec\_BV16(sp)}$
%$\textit{pc\_high}$ holds the most significant 8 bits and $\textit{pc\_low}$
%holds the least significant 8 bits.}.
%
%
% \textbf{NMI} - Non-maskable interrupts cannot be disabled
% by the programmer. Then, when a device makes a request, $sp$ is pushed,
% $pc$ receives $66H$ (102 in decimal), $\textit{iff1}$ is reset, $\textit{iff2}$ stores
% $\textit{iff1}$ and the refresh register is updated.
%  
%\begin{sloppypar}
%\bf updateStack\rm (\rm \{ \rm (\it sp\_minus\_two  $\mapsto$  \it pc\_low\rm )\rm ,\rm (\it sp\_minus\_one  $\mapsto$ \it pc\_high \rm ) \rm \}\rm )$\para$
%
%\it sp \rm := \it sp\_minus\_two  $\para$ \it pc \rm := \rm 1\rm 0\rm 2 $\para$ \it iff1\rm :=\rm 0  $\para$  \it iff2\rm := \it iff1 $\para$
%
%\it r\_ \rm := \it update\_refresh\_reg\rm (\it r\_\rm )\\
%\end{sloppypar}
%
%  \textbf{INT} - Maskable Interrupt is usually reserved for important functions
%  that can be enabled and disabled by the programmer. When a maskable interrupt
%  action happens, both $\textit{iff1}$ and $\textit{iff2}$ are cleared,
%  disabling the interrupts, $sp$ is pushed, the refresh register is updated and
%  the other effects depend on the interrupt mode register (\it im\rm).
% 
%
% \begin{itemize}
%   
%  \item The mode 0 is compatible with 8080 and this mode is selected when \it im
%  \rm = \rm ( \rm 0 $\mapsto$  \rm 0 \rm ). When a  non-maskable interrupt
%  happens, the CPU fetches an instruction of one byte from an external device,
%  usually an RST instruction, and the CPU executes it.
%  The instruction code is received from an external device by data bus and it is
%  represented by integer parameter called $\textit{byte\_bus}$.
%
% 
%  \item The mode 1 is the simplest and this mode is selected when \it im \rm =
%  \rm ( \rm 0  $\mapsto$  \rm 1 \rm ). Simply, when a non-maskable interruption
%  happens, the program counter receives $38H$ (56 in decimal).
%  
%  \item The mode 2 is the most flexible and this mode is selected when \it im
%  \rm = \rm ( \rm 1  $\mapsto$ \rm 1 \rm ). When a non-maskable interruption
%  happens, an indirect call can be made to any address memory. The program
%  counter receives a bit vector of size 16 composed with two parts: the most
%  significant part of the $\textit{i\_}$ register and the least significant part
%  of the $\textit{byte\_bus}$ with the last bit cleared.
%  
% \end{itemize}
%
